draft 1

### 2015-08-04 final write up/draft.rmd
draft update

Project decision makers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% (Flyvbjerg, 2006). A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average (Flyvbjerg and Budzier, 2011). Complex longer term projects are necessary to produce  railways, houses, websites, and movies which involve professionals across a many industries. Depending on the industry, consulting companies are enlisted to complete portions of a project, and some of the financial risk of the overall project may be transferred to the consultant. This is achieved by requiring a fixed flat fee from the consultant, meaning they have committed to perform their role for the nominated price. In a similar way to infrastructure and IT projects, consultants are susceptible to underestimating the amount of their time that will be spent in their role, and must wear the cost of extra employee hours spent. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects in a private business are multifold. Employee morale deteoriates and employees may view themselves as incompetent if they are responsible for the project's delivery. Employees also experience stress attempting to complete the project within a disappearing budget and may perform lower quality work in exchange for speed. This is unfortunate if the reality is that the project fee was under estimated. Unprofitable projects also limit a business' opportunity to invest money into marketing or staff training, which would improve a business' standing. The financial risk taken on fixed flat fee projects (and disappointing results) may also discourage business' from taking different risks with better potential pay-offs, such as expanding the business into a new area. Clearly, failure to predict fixed fee project costs can negatively affect a business' growth opportunities and staff morale.

@@

The challenge for consultants to reduce the risk of a project going over budget, in terms of hours spent, is not easily solved. If a job does go over budget, the business experiences reduced profits or  even losses as they are contractually bound to complete the work. The reason this task is challenging is that complex projects are always different from each other, and similar work may take significantly different amounts of time. For example, an ad filmed in a remote location for chocolate will cost significantly more than an ad filmed on an exisitng set, even though the product structure is almost identical. A fixed product price, which you could apply to repetive sales such as appliances, is inappropriate. One option is charging for consultants' time by the hour; a zero risk fee structure. However in many industries this is not competitive enough. Other consultancies may be willing to offer a fixed flat fee for the same work, reducing financial risk for the client, and luring the client away. Another fee strategy may be to track worker hours, and stop work once the consulting fee has been depleted. Then, negotiate further fees, or variations, with the client and persuade them to accept. This also satisfactorily reduces consultant financial risk, and in some cases, depending on the nature of the work or uncontrollable changes, this method can be effective. However, if the project has been following a predictable process, consultants often prefer to wear losses or low profits to preserve the relationship with their client. Consultant's may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry over creating financial friction. Clearly, there is no straightforward solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In some indusries, fixed flat fee structures are expected and business' must conform to survive.

@@

@@

- research has gone into targeting common repeated points in a projects timeline where things can go awry. reference
- example: find research, IT projects? nothing in engineering or film?
- Although no solutions are available, the use of reference class forecasting is being pioneered in england for large government infrastructure
- method for fee pricing has remained the same for decades - detailed analysis of project, add up time based on tasks, allow buffer time for error.

Reference class forecasting
* predicting cost of construction from other perspectives - builders.

Other industries
* predicition of building construction costs - statistical analysis

>>>>

Project decision makers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% (Flyvbjerg, 2006). A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average (Flyvbjerg and Budzier, 2011). These statistics cast a concerning glimpse at the financial story playing out behind the scenes of these projects which can cause job losses, ruin private businesses, and burn through government balancesheets. Complex long term projects are necessary to produce  railways, houses, websites, and movies. Depending on the industry, the the financial risk of completing a project on budget is transferred to consulting companies who are enlisted to complete portions of a project. The risk is transferred by requiring a fixed flat fee from the consultant, meaning they commit to complete their role for a nominated price. In a similar way to infrastructure and IT projects, consultants are susceptible to underestimating the amount of their time that will be spent in their role, and must wear the cost of extra employee hours spent. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects in a private business are multifold. Employee morale deteoriates and employees may view themselves as incompetent if they are responsible for the project's delivery. Employees also experience stress attempting to complete the project within a disappearing budget and may perform lower quality work in exchange for speed. This is unfortunate if in reality, the project fee was under estimated. Unprofitable projects also limit a business' opportunity to invest money into marketing or staff training, which would improve a business' standing. The financial risk taken on fixed flat fee projects may also discourage business' from taking different risks with better potential pay-offs, such as expanding the business into a new area. Clearly, predicting fixed fee project costs more accurately could positively affect a business' growth opportunities and staff morale.

@@

The challenge for consultants to reduce the risk of a project going over budget, in terms of hours spent, is not easily solved. The reason is that complex projects are always different from each other, and similar work may take significantly different amounts of time. For example, an ad filmed in a remote location for chocolate will cost significantly more than an ad filmed on an exisitng set, even though the product structure is almost identical. Therefore, a fixed product price, which you could apply to products such as appliances, is inappropriate for film projects. One option is charging for consultants' time and the hiring of materials by the hour; a zero risk fee structure. However in many industries this is not competitive enough. Other consultancies may be willing to offer a fixed flat fee for the same work, reducing financial risk for the client, and luring the client away. Another fee strategy may be to track worker hours, and stop work once the consulting fee has been depleted. Then, negotiate further fees, or variations, with the client and persuade them to accept. This also satisfactorily reduces consultant financial risk, and in some cases, depending on the nature of the work or uncontrollable changes, this method can be effective. However, if the project has been following a predictable process but the hours are nevertheless inflating beyond expectation (a common situation), consultants may prefer to wear losses or low profits to preserve the relationship with their client. Consultant's may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry over creating financial friction. Clearly, there is no straightforward solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In some indusries, fixed flat fee structures are expected and business' must conform to survive.

@@

Little research has been done into surveying the pricing methods of consulting projects, however

- current practise: literature available on IT and construction projects
        - detailed analysis of project
        - comparison to previous projects by experts - case based reasoning
        - intuition
        - some predictive statistics based on past data 
        - reviewing - trying to find places where things keep going wrong

- research: dozens of papers using statistical and machine learning to predict project costs
        - construc
        - IT
        - film costs?

- has any of this research been mobilised into practise, if not, why?
        - reference class forecasting - pioneered in british government large gov infrastructure

@@

- method for fee pricing has remained the same for decades - detailed analysis of project, add up time based on tasks, allow buffer time for error.
